Design Document: Monte Carlo Maximum Likelihood for Generalized Linear Mixed Models

### 2014-11-20 NewDesignDoc/DesignDocGLMM.tex
Updating design doc MCLA derivs
Erroneously did not take into account that ftwiddle depended on theta. Thus, terms were missing.

Since $u$ is unobservable, the log likelihood must be expressed by integrating out the random effects:
\begin{align}
l(\theta)=\log \int f_\theta(y|u) f_\theta(u) \; du
\end{align}
For most datasets, this integral is intractible. In these cases, performing even basic inference on the likelihood is not possible. Instead of evaluating the integral, \citet{geyer:thom:1992} suggest using a Monte Carlo approximation to the likelihood. Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.  Then the Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}
\end{align}
the gradient vector of the MCLA is
\begin{align}
\nabla l_m(\theta)= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }, \label{eq:MCLAgradient}
\end{align}
and the Hessian matrix of the MCLA is
\begin{align}
\nabla^2 l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla^2 \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \nabla l_m(\theta) (\nabla l_m(\theta) )'  \\
&+\dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) (\nabla \log f_\theta(u_k,y))'   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }.
\end{align}

@@

>>>>

Since $u$ is unobservable, the log likelihood must be expressed by integrating out the random effects:
\begin{align}
l(\theta)=\log \int f_\theta(y|u) f_\theta(u) \; du
\end{align}
For most datasets, this integral is intractible. In these cases, performing even basic inference on the likelihood is not possible. Instead of evaluating the integral, \citet{geyer:thom:1992} suggest using a Monte Carlo approximation to the likelihood. Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.  Then the Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}.
\end{align}
When $\tilde{f}$ depends on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
\begin{align}
\nabla l_m(\theta)= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f} (u_k)  \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }, \label{eq:MCLAgradient}
\end{align}
and the Hessian matrix of the MCLA is
\begin{align}
\nabla^2 l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla^2 \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \nabla l_m(\theta) (\nabla l_m(\theta) )'  \\
&+\dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) (\nabla \log f_\theta(u_k,y))'   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }.
\end{align}

@@

\appendix
\section{MCLA calculations}
The Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}.
\end{align}

\subsection{MCLA derivatives when $\tilde{f}$ depends on $\theta$}
When $\tilde{f}$ contains $\theta$, the gradient of the MCLA is
\begin{align}
\nabla l_m(\theta) &= \dfrac{ \nabla \left[ \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}  \right] }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \\
&= \dfrac{   \sum_{k=1}^m  \dfrac{\nabla f_\theta(y,u_k)   }{\tilde{f}(u_k)} - \dfrac{ f_\theta(y,u_k)  \nabla \tilde{f}(u_k) }{\left(\tilde{f}(u_k) \right)^2 } }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \\
&= \dfrac{   \sum_{k=1}^m  \dfrac{\nabla f_\theta(y,u_k)}{f_\theta(y,u_k)} \dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)} -
 \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)} \dfrac{ \nabla \tilde{f}(u_k) }{\tilde{f}(u_k) } }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \\
&= \dfrac{   \sum_{k=1}^m  \nabla \log f_\theta(y,u_k) \dfrac{f_\theta(y,u_k)}{\tilde{f}(u_k)} -
 \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  \nabla \log \tilde{f}(u_k)  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \\
&= \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} 
\end{align}

Then the Hessian of the MCLA is
\begin{align}
\nabla^2 l_m (\theta) &= \nabla \left[ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}} \right] \\
&= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k)   \right]  \nabla \left[ \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)} \right]  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
&-\left[\dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{ \left( \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)} \right)^2} \right] \nabla \left[ \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)} \right]\\
&= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k)   \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
&-\left[\dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -  \nabla \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{ \left( \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)} \right)^2} \right] \left[  \nabla \log f_\theta(y,u_k)  -  \nabla \log \tilde{f}(u_k)     \right]'\\
&= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k)   \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
&- \left[ \nabla l_m(\theta)  \right] \left[ \nabla l_m(\theta)  \right]'
\end{align}

\subsection{MCLA derivatives when $\tilde{f}$ independent of $\theta$}

When $\tilde{f}$ is independent of $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
\begin{align}
\nabla l_m(\theta)= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y)  \dfrac{f_\theta(u_k,y)   }{\tilde{f}(u_k)}\right) }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) } \label{eq:MCLAgradient}
\end{align}
and the Hessian matrix of the MCLA is
\begin{align}
\nabla^2 l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla^2 \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }- \nabla l_m(\theta) (\nabla l_m(\theta) )'  \\
&+\dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) (\nabla \log f_\theta(u_k,y))'   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right)}{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }.
\end{align}
